Collective excitations in competing phases in two and three dimensions

### 2024-03-26 paper/main.tex
results

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/afm_cdw_peaks_details.pdf}
    \caption{\red{Muss es im log-Argument nicht $U_0-U$ heissen, um klar zu sein?} 
		The upper panels (a) show double logarithmic plots of the positions $\omega_0$ 
		of the peak in $\spectral{l.AFM}$ relative to the lower edge of the two-particle continuum $\omega_-$.
    The lower panels (b) depict the respective weights $W_0$.
    Again, we choose $V=1.2t$ for the square lattice (left column) and 
		$V=0.8t$ for the simple cubic lattice (right column), respectively.
    The interaction $U_0$ denotes the highest value of $U$ for which 
		there is no peak below the two-particle continuum, i.e., for $U> U_0$ a peak occurs.
		\red{Ja, was gilt? Wann gibt's einen Peak, wann nicht? Bitte konsistent im ganzen Artikel bezeichnen.}
    For the square lattice, we find $U_0 \approx 3.372t$ and for the simple cubic lattice $U_0 \approx 4.479t$.
    The parameter range depicted here places the system within the CDW phase.
    The fits are of the same kind as in Fig.\ \ref{fig:sc_in_cdw_behavior}.}
    \label{fig:afm_cdw_peaks_details}
\end{figure*}

@@

Finally, we investigate the same spectral functions close to the AFM-CDW phase transition
as depicted in Fig.\ \ref{fig:resolvent_overview_AFM}.
We choose $V=1.2t$ for the square lattice and $V=0.8t$ for the simple cubic lattice so that
the phase transition is located at $U=4.8t$ in both cases. 
The panels (a) and (b) show the system close to the phase transition at $U=4.8t \pm 0.05t$.
Again, the SC-related spectral functions are identical in the CDW and the AFM phases
due to the absence of a finite SC order.
Within the continuum, their weight is shifted towards the upper edge.
On the square lattice only, there is a peak above the continuum.
It is likely to be smeared out if lifetime effects were included.
In the AFM phase, both $\spectral{l.AFM}$ and $\spectral{CDW}$ have a peak close to, but yet below, the two-particle continuum. 
The peak of the former is at a lower energy than the peak of the latter.
In the CDW phase, this behavior is reversed, i.e., the peak of $\spectral{CDW}$ lies lower
than the one in $\spectral{l.AFM}$.
Analyzing the power-law behavior of the real part described above, see also
App.\ \ref{sec:fit_greens_functions}, we confirm that these peaks are $\delta$ peaks.
Moving further away from the phase transition, here $U=6.2t$ in panels (c) \blue{and $U=3.4t$ in panels(d)}
\red{Im Text war (d) benannt, aber das existiert ja nicht.}, 
the upper peak merges with the two-particle continuum while the lower one persists.
We emphasize that the spectral function of the transverse magnon $\spectral{t.AFM}$ always 
displays a $\delta'$ peak at $\omega=0$, independent of how far away the parameters are chosen
from the phase transition. This is the expected behavior of a Goldstone boson.

@@

With the prevalent peaks discussed, we turn towards their behavior as the system approaches the phase transition at $V=0$ for $U=-2.5t$.
First and foremost, we investigate the peak in $\spectral{Phase}$ and $\spectral{Higgs}$ in the CDW phase.
The evolution of its position $\omega_0$ and its weight $W_0$ as $V$ is varied can be viewed double logarithmically in \autoref{fig:sc_in_cdw_behavior}.
For $V \ll t$, the position grows with the square root of $V$. This behavior then transitions to a purely linear growth for large $V$.
Dividing $\omega_0$ by the gap $\Delta_\text{CDW}$ results in a plateau for large $V$ while leaving the square root behavior at the beginning mostly unchanged.

The peak weights $W_0$ fall off as $1/\sqrt{V}$ for $V \ll t$. 
Afterward, it reaches a shallow minimum and rises slightly to reach a constant value.

Specifically, we learn that for $V \ll t$, the peak position follows $\omega_0 = \alpha \sqrt{V}$ and its weight follows $W_0 = \beta / \sqrt{V}$, 
where $\alpha$ and $\beta$ are some constants.
This allows us to write $W_0 = \beta / (\alpha \omega_0)$. 
Furthermore, we know that the spectral functions need to be perfectly antisymmetrical about $\omega = 0$, i.e., we can observe
\begin{align}
    \mathcal{A}_\text{Phase} (\omega \ll t) &= W_0 (\delta (\omega - \omega_0) - \delta (\omega + \omega_0)) \nonumber \\
        &= - \frac{2\beta}{\alpha} \frac{\delta (\omega + \omega_0) - \delta (\omega - \omega_0)}{2\omega_0} \,.
\end{align} 
Taking the limit of $V \to 0$ corresponds to the limit $\omega_0 \to 0$, i.e., leading to an expression reminiscent of symmetric derivative
\begin{equation}
    \lim_{h \to 0} \frac{f(x + h) - f(x - h)}{2h} = f'(x)\,,
\end{equation}
where the equality holds for any differentiable $f$.
Applying this to our case yields
\begin{align}
    \lim_{\omega_0 \to 0} \mathcal{A}_\text{Phase} (\omega \ll t) = - \frac{2 \beta}{\alpha} \delta'(\omega)\,.
\end{align}
This analysis reinforces our previous identification of the phase peak as $\delta' (\omega)$.

@@

As a next step, we want to analyze the peak in $\spectral{CDW}$ that occurs if the system is in the SC phase ($V<0$).
A plot of its position $\omega_0$ and weight $W_0$ can be seen in \autoref{fig:sc_in_cdw_behavior}.
First and foremost, the upper panels (a) and (b) depict the peak positions and weights, respectively.
\textcolor{\markEdited}{For $|V| \ll t$, the behavior of both the weight and the position of the peak is the same as for the previously discussed peak in $\spectral{SC}$ within the CDW phase.}

\textcolor{\markEdited}{Moving towards larger $|V|$, the growth of the peak position slows down, while the weights start to drop off rapidly.
However, we are unable to compute the Green's functions for additional, even larger $|V|$ because the matrix $\mM$ is not nonnegative anymore after $V\approx -0.28t$ (square lattice) or $V\approx -0.34t$ (simple cubic lattice).
Note, that our proof for the nonnegativity is based on the system being in thermal equilibrium.}
Therefore, we conclude that beyond these values of $V$, the superconducting phase is not the true ground state of the system, see the red shading in \autoref{fig:phase_diagram}.
Specifically, the $s$-wave superconducting phase on the square lattice seems to be even smaller than first estimated using the $d_{x^2 -y^2}$ data from \cite{Micnas88b}.
In recent research, there has been evidence for a phase-separated state for negative values of $V$ on a square lattice, which would explain our results here \cite{Linner23}.
It would be reasonable to assume, that this phase occurs in both the 2- and 3-dimensional case.
The magnitude of $V$ for which our calculation stops working is in the same order as in the referenced paper.
Note, however, that we did our computations at $T=0$ while the reference worked on finite temperature systems.

@@

Lastly, we investigate how the modes behave as the system transitions between the CDW and AFM phases.
Similar to the previous subsection, we plot the occurring peaks in $\spectral{CDW}$ and $\spectral{l.AFM}$ within the respective phase in \autoref{fig:afm_cdw_peaks_overview}.
On the right side of each panel, i.e., for $U > zV$, the system is in the AFM phase (green shading).
On the left side, it is in the CDW phase (orange shading).
In general, the peak in $\spectral{l.AFM}$ behaves exactly as the peak in $\spectral{CDW}$ if the phases are swapped.

The panel (a) shows the peak positions unmodified. They essentially move linearly as $U$ is varied.
It is noteworthy, that in the AFM phase, the peak in $\spectral{l.AFM}$ lies lower than the one in $\spectral{CDW}$.
Naturally, this is reversed, if the system is in the CDW phase.
This is to be expected, as creating an AFM excitation should be energetically more favorable if the system is already in said phase.

Additionally, we plot in (b) the peak positions relative to the lower edge of the two-particle continuum.
We see that the lower-lying peak, i.e., the peak corresponding to the phase the system is in, 
strays further away from the continuum as the system moves away from the phase boundary.
The other peak, however, moves ever closer to the aforementioned continuum until they reach it.
At this point, the weight of said peak also vanishes, see panel (c).
This occurs around $U \approx 4.8t \pm 1.428t$ for the square lattice and around $U \approx 4.8t \pm 0.321t$ for the simple cubic lattice.

On the square lattice, the weights of the peaks in $\spectral{l.AFM}$ in the AFM phase (the same applies to $\spectral{CDW}$ in the CDW phase) grow as the system moves away from the phase transition.
On the simple cubic lattice, however, the weights fall off very slightly.

In \autoref{fig:afm_cdw_peaks_details}, we revisit panels (b) and (c) from the previous figure.
We restrict the analysis to $\spectral{l.AFM}$ in the CDW phase as the reverse case yields precisely the same data.
This time around the scaling is double-logarithmic and we consider $U$ relative to those values at which we cannot find a peak in $\spectral{l.AFM}$.
This occurs around $U_0 = 3.372t$ on the square lattice and around $U_0 = 4.479t$ on the simple cubic lattice.
This yields that the peak positions $\omega_0$ (a) behave quadratically as they move into the two-particle continuum, i.e., $\omega_0 \propto (U-U_0)^2$,
while their weights $W_0$ (b) drop off linearly, i.e., $W_0 \propto (U-U_0)$.

>>>>

\begin{figure*}
    \centering
    \includegraphics[width=.98\textwidth]{plots/afm_cdw_peaks_details.pdf}
    \caption{\red{Muss es im log-Argument nicht $U_0-U$ heissen, um klar zu sein?}  \blue{Wir plotten $U\in(U_0, 4.8t)$, also $U_0 < U$}
		The upper panels (a) show double logarithmic plots of the positions $\omega_0$ 
		of the peak in $\spectral{l.AFM}$ relative to the lower edge of the two-particle continuum $\omega_-$.
    The lower panels (b) depict the respective weights $W_0$.
    Again, we choose $V=1.2t$ for the square lattice (left column) and 
		$V=0.8t$ for the simple cubic lattice (right column), respectively.
    The interaction $U_0$ denotes the highest value of $U$ for which 
		there is no peak below the two-particle continuum, i.e., for $U> U_0$ a peak occurs.
		\red{Ja, was gilt? Wann gibt's einen Peak, wann nicht? Bitte konsistent im ganzen Artikel bezeichnen.}
    For the square lattice, we find $U_0 \approx 3.372t$ and for the simple cubic lattice $U_0 \approx 4.479t$.
    The parameter range depicted here places the system within the CDW phase.
    The fits are of the same kind as in Fig.\ \ref{fig:sc_in_cdw_behavior}.}
    \label{fig:afm_cdw_peaks_details}
\end{figure*}

@@

Finally, we investigate the same spectral functions close to the AFM-CDW phase transition
as depicted in Fig.\ \ref{fig:resolvent_overview_AFM}.
We choose $V=1.2t$ for the square lattice and $V=0.8t$ for the simple cubic lattice so that
the phase transition is located at $U=4.8t$ in both cases. 
The panels (a) and (b) show the system close to the phase transition at $U=4.8t \pm 0.05t$.
Again, the SC-related spectral functions are identical in the CDW and the AFM phases
due to the absence of a finite SC order.
Within the continuum, their weight is shifted towards the upper edge.
On the square lattice only, there is a peak above the continuum.
It is likely to be smeared out if lifetime effects were included.
In the AFM phase, both $\spectral{l.AFM}$ and $\spectral{CDW}$ have a peak close to, but yet below, the two-particle continuum. 
The peak of the former is at a lower energy than the peak of the latter.
In the CDW phase, this behavior is reversed, i.e., the peak of $\spectral{CDW}$ lies lower
than the one in $\spectral{l.AFM}$.
Analyzing the power-law behavior of the real part described above, see also
App.\ \ref{sec:fit_greens_functions}, we confirm that these peaks are $\delta$ peaks.
Moving further away from the phase transition, here $U=6.2t$ in panels (c) \blue{and $U=3.4t$ in panels (d)}
\red{Im Text war (d) benannt, aber das existiert ja nicht.}, 
the upper peak merges with the two-particle continuum while the lower one persists.
We emphasize that the spectral function of the transverse magnon $\spectral{t.AFM}$ always 
displays a $\delta'$ peak at $\omega=0$, independent of how far away the parameters are chosen
from the phase transition. This is the expected behavior of a Goldstone boson.

@@

Having discussed the dominant peaks, we address their behavior when 
the system approaches the phase transition \blue{between the SC and CDW phase} at $V=0$ for $U=-2.5t$.
First, we investigate the identical peak in $\spectral{Phase}$ and $\spectral{Higgs}$, respectively, in the CDW phase.
The evolution of its position $\omega_0$ and its weight $W_0$ as a function of $V$ is depicted
double logarithmically in Fig.\ \ref{fig:sc_in_cdw_behavior}.
For $V \ll t$, the position increases with the square root of $V$. 
This behavior crosses over to a purely linear growth for large $V$.
The gap position $\omega_0$ relative to the gap $\Delta_\text{CDW}$ shows a plateau for large $V$ 
while leaving the square root behavior for small values of $V$ hardly altered.
The peak's weight $W_0$ decreases like $1/\sqrt{V}$ for $V \ll t$,
reaching a shallow minimum for intermediate valus of $V$, and 
eventually saturating at a constant value.

Summarizing, we see that for $V \ll t$ the peak position follows 
$\omega_0 = \alpha \sqrt{V}$ and its weight  $W_0 = \beta / \sqrt{V}$, 
where $\alpha$ and $\beta$ are some constants. Thus, we conclude $W_0 = \beta / (\alpha \omega_0)$. 
Furthermore, we know that the spectral functions are antisymmetric, i.e., we find
\bs
\begin{align}
    \mathcal{A}_\text{Phase} (\omega \ll t) &= W_0 (\delta (\omega - \omega_0) - \delta (\omega + \omega_0)) 
		\\
    &= - \frac{2\beta}{\alpha} \frac{\delta (\omega + \omega_0) - \delta (\omega - \omega_0)}{2\omega_0}.
\end{align} 
\es
Taking the limit $V \to 0$ corresponds to the limit $\omega_0 \to 0$, i.e., leading to a ratio tending to the
derivative
\begin{equation}
    \lim_{h \to 0} \frac{f(x + h) - f(x - h)}{2h} = f'(x).
\end{equation}
Applying this to the spectral peak yields
\begin{align}
    \lim_{\omega_0 \to 0} \mathcal{A}_\text{Phase} (\omega \ll t) = - \frac{2 \beta}{\alpha} \delta'(\omega),
\end{align}
confirming our previous analysis of the phase peak being proportional to  $\delta' (\omega)$.

@@

Next, we analyze the peak in $\spectral{CDW}$ occurring if the system is in the SC phase for $V<0$.
A plot of its position $\omega_0$ and weight $W_0$ is given in Fig.\ \ref{fig:cdw_in_sc_behavior}.
The upper panels (a) and (b) depict the peak position and weight, respectively.
For $|V| \ll t$, the behavior of both of them is the same as for the 
previously discussed peak in $\spectral{SC}$ in the CDW phase.
Increasing $|V|$, the increase of the peak position slows down while the weights start to drop rapidly.
We cannot compute the Green's functions for even larger $|V|$ with $V<0$ because the dynamical matrix $\mM$ 
acquires negative eigenvalues below $V\approx -0.28t$ (square lattice) or $V\approx -0.34t$ (simple cubic lattice). 
We recall that this indicates that the assumed phase, i.e., the SC phase is not stable. 
Therefore, we conclude that for sufficiently negative values of $V$, 
the superconducting phase is not the true ground state as indicated by the red shading in Fig.\ \ref{fig:phase_diagram}.
Specifically, the region of the SC phase on the square lattice 
appears to be even smaller than shown by the $d_{x^2 -y^2}$ data from Ref.\ \cite{Micnas88b}.
Recently, there has been evidence for phase separation for negative values of $V$ 
on the square lattice, at least at finite temperature \cite{Linner23}. 
Thus, we interpret the negative eigenvalues in the dynamical matrix as indicators of phase separation. 
Physically, a strong attractive interaction clearly favors phase separation since the particles prefer to stay close together 
so it is highly plausible that they gather in one region of the sample while leaving the remainder essentially empty.
\red{Was ist in 3D - gibt es da nicht auch negative Eigenwerte usw.? Das solten wir dann hier
ebenfalls besprechen. Wo wollten wir denn die Diskussion des Tricks von Biermann publizieren?}

@@

Last, we investigate how the modes behave as the system passes from the CDW to the AFM phase.
As in the previous subsection, we plot the occurring peaks in $\spectral{CDW}$ and 
$\spectral{l.AFM}$ in the respective phases in Fig.\ \ref{fig:afm_cdw_peaks_overview}.
On the right-hand side of each panel, i.e., for $U > zV$, the system is in the AFM phase (green shading);
on the left-hand side, it is in the CDW phase (orange shading).
Qualitatively, the peak in $\spectral{l.AFM}$ behaves like the peak in 
$\spectral{CDW}$ if the phases are swapped. \red{Ist das exakt oder nur ungefaehr so?} \blue{tatsächlich exakt.}

Panel (a) shows the peak positions which shift essentially linearly upon varying $U$.
It is noteworthy, that in the AFM phase the peak in $\spectral{l.AFM}$ lies energetically
lower than the one \red{Welchen meinst Du? CDW oder l.AFM? Uneindeutig ...} in $\spectral{CDW}$. \blue{Der AFM-peak ist in der AFM-phase bei kleinerer Energie als der CDW-peak in der AFM-phase.}
This is reversed if the system is in the CDW phase.
We conclude that it is energetically more favorable to create an AFM excitation (magnon) than a CDW
exciton if the system is in the AFM phase. In the CDW phase, the creation of the CDW exciton is
energetically cheaper than the creation of a magnon. \red{Habe ich das so richtig dargestellt?} \blue{Ja.}

Additionally, we plot the peak positions in panel (b) relative to the lower edge of the two-particle continuum.
We see that the lower-lying peak, i.e., the peak corresponding to the phase in which the system is, 
shifts further away from the continuum as the system moves away from the phase boundary.
The other peak, however, shifts closer to the continuum until it dives into it.
At this point, the weight of the merging peak vanishes, see panels (c).
\blue{This occurs around $U \approx 4.8t + 1.428t$ ($\spectral{CDW}$) and $U \approx 4.8t - 1.428t$ ($\spectral{l.AFM}$) for the square lattice 
and around $U \approx 4.8t + 0.321t$  ($\spectral{CDW}$) and $U \approx 4.8t - 0.321t$  ($\spectral{l.AFM}$) for the simple cubic lattice.}
\red{Was willst Du mit dem $\pm$ ausdruecken? Verstehe ich nicht ... Besser etwas ausfuehrlicher trotz
einer gewissen Redundanz.}

The weights of the peaks in $\spectral{l.AFM}$ in the AFM phase (the same applies to $\spectral{CDW}$ in the CDW phase) grow as the system moves away from the phase transition.
The same applies to the peaks in $\spectral{CDW}$ in the CDW phase.

In Fig.\ \ref{fig:afm_cdw_peaks_details}, we revisit the data of panels (b) and (c) of the previous figure
on a double logarithmic scale.
We restrict the analysis to $\spectral{l.AFM}$ in the CDW phase as the swapped case of $\spectral{CDW}$ in the 
AFM phase shows the same behavior.
We consider $U$ relative to the value $U_0$ at which the peak in $\spectral{l.AFM}$ ceases to exist.
\red{Stimmt die Formulierung?} \blue{Ja.}
This occurs at $U_0 = 3.372t$ on the square lattice and at $U_0 = 4.479t$ on the simple cubic lattice.
The peak positions $\omega_0$ in panels (a) behave quadratically as they merge with the 
two-particle continuum, i.e., 
$\omega_0 \propto (U-U_0)^2$ while the weights $W_0$ in panel (b) decrease linearly, i.e., 
$W_0 \propto (U-U_0)$ \red{Stimmt das Vorzeichen?}. This behavior is generic for the peaks of bound states merging with
their continua at vanishing binding energy \cite{uhrig96b,uhrig96be,zhito13}.
